An Open-Source Implementation of the CMPS Algorithm for Assessing Similarity of Bullets

### 2022-03-24 submissions/resubmission-ju-hofmann/ju-hofmann.Rmd
reorganize the folder

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(tidyverse)
library(bulletxtrctr)
library(kableExtra)
library(patchwork)
library(ggpubr)

# source("../code/func_collection.R")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "\\textwidth",
  fig.align = "center",
#  cache=TRUE,
  dpi = 144,
  warning = FALSE,
  message = FALSE
)
```

@@

@@

In this paper we present an open-source implementation of the algorithm of the Congruent Matching Profile Segments (CMPS) method.
@cmps developed the CMPS method for "objective comparison of striated tool marks" and demonstrated its use in some examples of comparing bullet signature correlations.
Although @cmps conceptually describe the CMPS algorithm in their paper, the authors did not release an implementation of their method.
Thus, our effort here is to introduce the CMPS method to the `R` [@R] community and provide an open-source, publicly available implementation of the algorithm to use, review, and improve.
<!-- Our implementation is made available as part of the `R` package \pkg{cmpsR} on GitHub at \url{https://github.com/willju-wangqian/cmpsR}. -->
\wj{change}
Our implementation is made available as part of the `R` package \pkg{cmpsR} on CRAN.

<!-- https://s3-us-gov-west-1.amazonaws.com/cg-d4b776d0-d898-4153-90c8-8336f86bdfec/CIUS/downloads/2020/expanded-homicide-2020.zip -->
<!-- Firearm examination has  played an important role in  convictions [@fbiucr] in the United States criminal justice system. \hh{Do we have a number or a percentage of cases where firearm examination has played a role? or maybe gun violence? XXX} --> <!-- https://ucr.fbi.gov/crime-in-the-u.s/2019/crime-in-the-u.s.-2019/topic-pages/expanded-homicide --> 
According to the Uniform Crime Reporting Program of FBI (Federal Bureau of Investigation), "more than 76 percent (76.7) of the homicides for which the FBI received weapons data in 2020 involved the use of firearms" [@fbiucr]. At the same time, the number of murder victims jumped between 2019 and 2020 by more than 23 percent (23.4) to 17754 [@fbiucr20]. This increase is unprecedented  and highlights the important role that  firearm examination plays in all these cases. 
<!-- homicides in 2019: 14391; in 2020: 17754 -->
An important task of firearm examination is to answer the question of whether two pieces of evidence come from the same source or whether a piece of evidence matches a sample obtained from a specific firearm.
Here, in particular, we are interested to determine whether two bullets were fired from the same gun barrel.
Assessing the similarity between two bullets is based on a comparison of striation marks acquired during the firing process as bullets are propelled through the barrel.
Current state of art sees firearms examiners make an assessment of similarity based on a visual comparison, generally, using a comparison microscope [@afte].
This practice has been criticized for its lack of objectivity and the associated problem of determining valid error rates [@pcast].

@@

The datasets we worked with come from the James Hamby Consecutively Rifled Ruger Barrel Study [@brundage; @hamby; @Hamby:2019], in particular, Hamby set 252 and Hamby set 44.
For each Hamby set, a total of 35 bullets is fired from (the same) ten consecutively manufactured Ruger P-85 pistol barrels.
Two bullets are fired from each barrel, making up a set of 20 reference bullets.
An additional 15 bullets are fired from these ten barrels in a fashion unknown to the study participant.
The aim of the Hamby Study was to have firearms examiners identify which barrel each of the 15 questioned bullets was fired from.
The Ruger P-85 barrels are traditionally rifled barrels with six grooves and lands as shown in \autoref{fig:bullet}.
During the firing process, grooves and lands are engraved on a bullet.
Firearms examiners use striation marks on land engraved areas (LEAs) for their visual comparison.
For algorithmic purposes, 3D topographical images of land engraved areas were obtained and stored in x3p format (XML 3-D Surface Profile).
The x3p format provides a standard way of exchanging 2D and 3D profile data.
It conforms to the ISO5436-2 standard [<http://sourceforge.net/p/open-gps/mwiki/X3p/>] adopted by the OpenFMC (Open Forensic Metrology Consortium), a group of firearm forensics researchers who contributes to the establishment of best practices of using metrology in forensic science.
Hamby set 252 was scanned using a NanoFocus lens at 20x magnification with the scan resolution being 1.5625 \textmu m $\times$ 1.5625 \textmu m per pixel.
Hamby set 44 was scanned at the Roy J Carver High-Resolution microscopy lab at Iowa State.
These scans were acquired with a Sensofar Confocal Light Microscope at 20x magnification for a nominal resolution of 0.645 \textmu m $\times$ 0.645 \textmu m per pixel.
Both Hamby set 252 and Hamby set 44 are publicly available from the NIST Ballistics Database Project [@nistdb].

@@

The thin white horizontal line in \autoref{fig:process}(a) indicates which cross-section was identified by the algorithm to represent the LEA; \autoref{fig:process}(b) shows the corresponding cross-sectional view.
Groove engraved areas are removed from the analysis as indicated by the vertical blue lines in \autoref{fig:process}(c) and \autoref{fig:process}(d)).
A non-parametric LOESS smooth [@loess] is fitted to capture the bullet curvature (\autoref{fig:process}(e)) and, finally, the **bullet signature** (\autoref{fig:process}(f)) is obtained as residuals of the cross-section and the fitted smooth.
Note, that in @cmps bullet signatures are referred to as bullet profiles.
However, to avoid confusion, we distinguish the notion of bullet signatures from bullet profiles.
Bullet profiles are shown in panels (b), (c), and (d) of \autoref{fig:process}, while \autoref{fig:process}(f) shows the corresponding bullet signature.
Identifying the groove engraved areas correctly is fundamental for a correct down-stream analysis of the signatures.
To allow for a human-in-the middle inspection and intervention we have provided an interactive web-application, implemented as `R` Shiny App [@shiny], \pkg{bulletinspectR} to identify and correct those errors.
An example of the extraction process with corresponding code and parameter settings can be found on Github at \url{https://github.com/willju-wangqian/CMPSpaper/tree/main/reproducible}. Note that the process of extracting signatures might be different from the one used in @cmps because no code or parameter settings are made available publicly.
\wj{can be found in the supplementary materials of this paper? }

@@

Most algorithms for comparing striation marks are based on the digitized signatures and produce a similarity score [@song2005, @ChumbleyL_Scott2010VoTM, @aoas, @pmid30444940].
The congruent matching profile segments (CMPS) algorithm, developed by @cmps for "objective comparison of striated tool marks", is one such algorithm.
The algorithm's main idea is to take a set of consecutive and non-overlapping basis segments from the comparison and for each segment find the "best" registration position on the reference (the other bullet signature) with respect to their cross-correlation values.
From a comparison of these registration positions, a **congruent registration position** is identified, and the number of basis segments taking the congruent registration position is the CMPS score.
Note that researchers in @cmps refer to the origin of basis segments as the reference, but in this paper we refer to it as the comparison signature.
High CMPS scores are achieved between more similar signatures and are therefore indicative of a same-source pair.
Low scores between pairs of signatures are attributed to different source pairs.
However, a specific threshold of the CMPS score to distinguish between same-source and different-source comparisons is not provided in @cmps, instead the threshold depends on the underlying structure of the data and the choice of parameters.   In a legal setting this variability is problematic because it allows for situations in which experts could choose parameters based on whether they are witnesses for the defense or the prosecution. 
Further research is needed to understand how to determine optimal threshold settings.
The CMPS algorithm can assist firearm examiners with drawing a conclusion about the source of a comparison pair.
Thus, in this paper we present an open-source implementation of the CMPS algorithm in the R package \pkg{cmpsR} available from both CRAN and Github \url{https://github.com/willju-wangqian/cmpsR}. This publicly available implementation calculates the CMPS score of a comparison using the following code:
\wj{ignore github?}

@@

<!--For more details, please check the documentation of `extract_feature_cmps` or use `?extract_feature_cmps`.-->

<!-- [examples https://journal.r-project.org/archive/2019/RJ-2019-044/RJ-2019-044.pdf] -->

<!-- [https://journal.r-project.org/archive/2019/RJ-2019-002/RJ-2019-002.pdf] -->

@@

<!-- ```{r hamby252-data-seg, eval=TRUE, echo=FALSE} -->
<!-- # for hamby 252 -->
<!-- N <- 8 # the number of different sets of parameters -->
<!-- CMPS_hamby252_results <- list() # a container for everything -->

<!-- #### setup span1 -->
<!-- CMPS_hamby252_results$span1 <- as.list(rep(0.25, N)) -->

<!-- #### setup signature name -->
<!-- CMPS_hamby252_results$signame <- as.list(rep("sigs25", N)) -->
<!--   # list("sigs75", "sigs25", "sigs25_1062", "sigs15") -->

<!-- #### setup npeaks.set -->
<!-- CMPS_hamby252_results$npeaks.set <- lapply(1:N, function(t) c(5,3,1)) -->

<!-- #### setup seg_length -->
<!-- CMPS_hamby252_results$seg_length <- as.list(c(25, 50, 75, 100, 125, 150, 175, 200)) -->

<!-- #### setup Tx -->
<!-- CMPS_hamby252_results$Tx <- as.list(rep(25, N)) -->

<!-- #### setup outlength -->
<!-- CMPS_hamby252_results$outlength <- vector(mode = "list", length = N) -->

<!-- #### create a plot title and a file name for each set of parameters -->
<!-- CMPS_hamby252_results$titlee <- list() -->
<!-- CMPS_hamby252_results$filename <- list() -->
<!-- for (i in 1:N) { -->
<!--   CMPS_hamby252_results$titlee[[i]] <- -->
<!--     paste0( -->
<!--       "npeaks.set=c(", -->
<!--       paste(CMPS_hamby252_results$npeaks.set[[i]], collapse = ","), -->
<!--       ")", -->
<!--       ", len=", -->
<!--       CMPS_hamby252_results$seg_length[[i]], -->
<!--       ", Tx=", -->
<!--       CMPS_hamby252_results$Tx[[i]], -->
<!--       ", \nspan1=", -->
<!--       CMPS_hamby252_results$span1[[i]], -->
<!--       ", span2=0.03" -->
<!--     ) -->
<!--   CMPS_hamby252_results$filename[[i]] <-  -->
<!--     paste( -->
<!--       "hamby252_segment", -->
<!--       CMPS_hamby252_results$span1[[i]]*100, -->
<!--       paste(CMPS_hamby252_results$npeaks.set[[i]], collapse = "-"), -->
<!--       CMPS_hamby252_results$seg_length[[i]], -->
<!--       CMPS_hamby252_results$Tx[[i]], -->
<!--       sep = "_" -->
<!--     ) -->
<!-- } -->

<!-- #### container for the cmps results and corresponding plots -->
<!-- CMPS_hamby252_results$cmps.table <- list() -->
<!-- CMPS_hamby252_results$plot <- list() -->

<!-- #### a global title for Hamby252 plots -->
<!-- com.title252 <- expression(paste( -->
<!--   "Hamby 252 - ", CMPS[max], " and ", bar(CMPS)[max], " Distribution" -->
<!-- )) -->

<!-- for(i in 1:N){ -->
<!--   CMPS_hamby252_results$cmps.table[[i]] <-  -->
<!--   read.csv(file = paste("./data/hamby252/",  -->
<!--                         CMPS_hamby252_results$filename[[i]],  -->
<!--                         ".csv", -->
<!--                         sep = "")) -->
<!-- } -->

<!-- CMPS_hamby252_results_seg <- CMPS_hamby252_results -->

<!-- ``` -->

<!-- ```{r hamby252-data-npeak, eval=TRUE, echo=FALSE} -->
<!-- N <- 8 # the number of different sets of parameters -->
<!-- CMPS_hamby252_results <- list() # a container for everything -->

<!-- #### setup span1 -->
<!-- CMPS_hamby252_results$span1 <- as.list(rep(0.25, N)) -->

<!-- #### setup signature name -->
<!-- CMPS_hamby252_results$signame <- as.list(rep("sigs25", N)) -->
<!-- # list("sigs75", "sigs25", "sigs25_1062", "sigs15") -->

<!-- #### setup npeaks.set -->
<!-- CMPS_hamby252_results$npeaks.set <-  -->
<!--   list(c(5), -->
<!--        c(5,2), -->
<!--        c(5,3,1), -->
<!--        c(10,6,2), -->
<!--        c(6,4,2,1), -->
<!--        c(10,7,4,2), -->
<!--        c(10,6,4,2,1), -->
<!--        c(10,8,6,4,2,1)) -->

<!-- #### setup seg_length -->
<!-- CMPS_hamby252_results$seg_length <-  -->
<!--   # as.list(c(25, 50, 75, 100, 125, 150, 175, 200)) -->
<!--   as.list(rep(50, N)) -->

<!-- #### setup Tx -->
<!-- CMPS_hamby252_results$Tx <- as.list(rep(25, N)) -->

<!-- #### setup outlength -->
<!-- CMPS_hamby252_results$outlength <- vector(mode = "list", length = N) -->
<!-- CMPS_hamby252_results$outlength[[7]] <- c(60, 90, 120, 150, 180) -->
<!-- CMPS_hamby252_results$outlength[[8]] <- c(50,75,100,125,150,200) -->

<!-- #### create a plot title and a file name for each set of parameters -->
<!-- CMPS_hamby252_results$titlee <- list() -->
<!-- CMPS_hamby252_results$filename <- list() -->
<!-- for (i in 1:N) { -->
<!--   CMPS_hamby252_results$titlee[[i]] <- -->
<!--     paste0( -->
<!--       "npeaks.set=c(", -->
<!--       paste(CMPS_hamby252_results$npeaks.set[[i]], collapse = ","), -->
<!--       ")", -->
<!--       ", len=", -->
<!--       CMPS_hamby252_results$seg_length[[i]], -->
<!--       ", Tx=", -->
<!--       CMPS_hamby252_results$Tx[[i]], -->
<!--       ", \nspan1=", -->
<!--       CMPS_hamby252_results$span1[[i]], -->
<!--       ", span2=0.03" -->
<!--     ) -->
<!--   CMPS_hamby252_results$filename[[i]] <-  -->
<!--     paste( -->
<!--       "hamby252_npeak", -->
<!--       CMPS_hamby252_results$span1[[i]]*100, -->
<!--       paste(CMPS_hamby252_results$npeaks.set[[i]], collapse = "-"), -->
<!--       CMPS_hamby252_results$seg_length[[i]], -->
<!--       CMPS_hamby252_results$Tx[[i]], -->
<!--       sep = "_" -->
<!--     ) -->
<!-- } -->

<!-- #### container for the cmps results and corresponding plots -->
<!-- CMPS_hamby252_results$cmps.table <- list() -->
<!-- CMPS_hamby252_results$plot <- list() -->

<!-- #### a global title for Hamby252 plots -->
<!-- com.title252 <- expression(paste( -->
<!--   "Hamby 252 - ", CMPS[max], " and ", bar(CMPS)[max], " Distribution" -->
<!-- )) -->

<!-- for(i in 1:N){ -->
<!--   CMPS_hamby252_results$cmps.table[[i]] <-  -->
<!--   read.csv(file = paste("./data/hamby252/",  -->
<!--                         CMPS_hamby252_results$filename[[i]],  -->
<!--                         ".csv", -->
<!--                         sep = "")) -->
<!-- } -->

<!-- CMPS_hamby252_results_npeak <- CMPS_hamby252_results -->

<!-- ``` -->

<!-- ```{r hamby44-data-seg, eval=TRUE, echo=FALSE} -->
<!-- N <- 8 # the number of different sets of parameters -->
<!-- CMPS_hamby44_results <- list() # a container for everything -->

<!-- #### setup span1 -->
<!-- CMPS_hamby44_results$span1 <- as.list(rep(0.25, N)) -->

<!-- #### setup signature name -->
<!-- CMPS_hamby44_results$signame <- -->
<!--   as.list(rep("sigs25_531", N)) -->

<!-- #### setup neapks.set -->
<!-- CMPS_hamby44_results$npeaks.set <- -->
<!--   lapply(1:N, function(t) c(5,3,1)) -->

<!-- #### setup seg_length -->
<!-- CMPS_hamby44_results$seg_length <- as.list(c(30, 45, 61, 90, 122, 150, 180, 210)) -->

<!-- #### setup outlength -->
<!-- CMPS_hamby44_results$outlength <- vector(mode = "list", length = N) -->

<!-- #### setup Tx -->
<!-- CMPS_hamby44_results$Tx <- as.list(rep(30, N)) -->

<!-- #### create a plot title and a file name for each set of parameters -->
<!-- CMPS_hamby44_results$titlee <- list() -->
<!-- CMPS_hamby44_results$filename <- list() -->
<!-- for (i in 1:N) { -->
<!--   CMPS_hamby44_results$titlee[[i]] <- -->
<!--     paste0( -->
<!--       "npeaks.set=c(", -->
<!--       paste(CMPS_hamby44_results$npeaks.set[[i]], collapse = ","), -->
<!--       ")", -->
<!--       ", len=", -->
<!--       CMPS_hamby44_results$seg_length[[i]], -->
<!--       ", Tx=", -->
<!--       CMPS_hamby44_results$Tx[[i]], -->
<!--       ", \nspan1=", -->
<!--       CMPS_hamby44_results$span1[[i]], -->
<!--       ", span2=0.03" -->
<!--     ) -->
<!--   CMPS_hamby44_results$filename[[i]] <-  -->
<!--     paste( -->
<!--       "hamby44_segment", -->
<!--       CMPS_hamby44_results$span1[[i]]*100, -->
<!--       paste(CMPS_hamby44_results$npeaks.set[[i]], collapse = "-"), -->
<!--       CMPS_hamby44_results$seg_length[[i]], -->
<!--       CMPS_hamby44_results$Tx[[i]], -->
<!--       sep = "_" -->
<!--     ) -->
<!-- } -->

<!-- #### container for the cmps results and corresponding plots -->
<!-- CMPS_hamby44_results$cmps.table <- list() -->
<!-- CMPS_hamby44_results$plot <- list() -->

<!-- #### a global title for Hamby44 plots -->
<!-- com.title44 <- expression(paste( -->
<!--   "Hamby 44 - ", CMPS[max], " and ", bar(CMPS)[max], " Distribution" -->
<!-- )) -->

<!-- for(i in 1:N){ -->
<!--   CMPS_hamby44_results$cmps.table[[i]] <-  -->
<!--   read.csv(file = paste("./data/hamby44/",  -->
<!--                         CMPS_hamby44_results$filename[[i]],  -->
<!--                         ".csv", -->
<!--                         sep = "")) -->
<!-- } -->

<!-- CMPS_hamby44_results_seg <- CMPS_hamby44_results -->

<!-- ``` -->

<!-- ```{r hamby44-data-npeak, eval=TRUE, echo=FALSE} -->
<!-- N <- 8 # the number of different sets of parameters -->
<!-- CMPS_hamby44_results <- list() # a container for everything -->

<!-- #### setup span1 -->
<!-- CMPS_hamby44_results$span1 <- as.list(rep(0.25, N)) -->

<!-- #### setup signature name -->
<!-- CMPS_hamby44_results$signame <- -->
<!--   as.list(rep("sigs25_531", N)) -->

<!-- #### setup neapks.set -->
<!-- CMPS_hamby44_results$npeaks.set <- -->
<!--   list(c(5), -->
<!--        c(5,2), -->
<!--        c(5,3,1), -->
<!--        c(10,6,2), -->
<!--        c(6,4,2,1), -->
<!--        c(10,7,4,2), -->
<!--        c(10,6,4,2,1), -->
<!--        c(10,8,6,4,2,1)) -->

<!-- #### setup seg_length -->
<!-- CMPS_hamby44_results$seg_length <- as.list(rep(61, N)) -->

<!-- #### setup outlength -->
<!-- CMPS_hamby44_results$outlength <- vector(mode = "list", length = N) -->
<!-- CMPS_hamby44_results$outlength[[7]] <- c(61, 91, 121, 151, 181) -->
<!-- CMPS_hamby44_results$outlength[[8]] <- c(61, 91, 121, 151, 181, 211) -->

<!-- #### setup Tx -->
<!-- CMPS_hamby44_results$Tx <- as.list(rep(30, N)) -->

<!-- #### create a plot title and a file name for each set of parameters -->
<!-- CMPS_hamby44_results$titlee <- list() -->
<!-- CMPS_hamby44_results$filename <- list() -->
<!-- for (i in 1:N) { -->
<!--   CMPS_hamby44_results$titlee[[i]] <- -->
<!--     paste0( -->
<!--       "npeaks.set=c(", -->
<!--       paste(CMPS_hamby44_results$npeaks.set[[i]], collapse = ","), -->
<!--       ")", -->
<!--       ", len=", -->
<!--       CMPS_hamby44_results$seg_length[[i]], -->
<!--       ", Tx=", -->
<!--       CMPS_hamby44_results$Tx[[i]], -->
<!--       ", \nspan1=", -->
<!--       CMPS_hamby44_results$span1[[i]], -->
<!--       ", span2=0.03" -->
<!--     ) -->
<!--   CMPS_hamby44_results$filename[[i]] <-  -->
<!--     paste( -->
<!--       "hamby44_npeak", -->
<!--       CMPS_hamby44_results$span1[[i]]*100, -->
<!--       paste(CMPS_hamby44_results$npeaks.set[[i]], collapse = "-"), -->
<!--       CMPS_hamby44_results$seg_length[[i]], -->
<!--       CMPS_hamby44_results$Tx[[i]], -->
<!--       sep = "_" -->
<!--     ) -->
<!-- } -->

<!-- #### container for the cmps results and corresponding plots -->
<!-- CMPS_hamby44_results$cmps.table <- list() -->
<!-- CMPS_hamby44_results$plot <- list() -->

<!-- #### a global title for Hamby44 plots -->
<!-- com.title44 <- expression(paste( -->
<!--   "Hamby 44 - ", CMPS[max], " and ", bar(CMPS)[max], " Distribution" -->
<!-- )) -->

<!-- for(i in 1:N){ -->
<!--   CMPS_hamby44_results$cmps.table[[i]] <-  -->
<!--   read.csv(file = paste("./data/hamby44/",  -->
<!--                         CMPS_hamby44_results$filename[[i]],  -->
<!--                         ".csv", -->
<!--                         sep = "")) -->
<!-- } -->

<!-- CMPS_hamby44_results_npeak <- CMPS_hamby44_results -->

<!-- ``` -->

<!-- ```{r rf-data-load, echo=FALSE} -->
<!-- rf.results <- list() -->
<!-- rf.results[[1]] <- read.csv(file = "./data/hamby252/hamby252_rf_results.csv") -->
<!-- rf.results[[2]] <- read.csv(file = "./data/hamby44/hamby44_rf_results.csv") -->
<!-- ``` -->

@@

The code generating all results of this manuscript can be found on Github at \url{https://github.com/willju-wangqian/CMPSpaper/tree/main/reproducible}
\wj{in the supplementary materials?}

@@

**Data**

@@

<!-- ```{r plot_thisisus} -->
<!-- rf.252.plot <- metric_plot_helper(rf.results[[1]], "rf.diff", scaled = TRUE, -->
<!--                                   subtitle = "random forest scores with metric diff") + -->
<!--   xlab(expression(RF[diff])) + -->
<!--   scale_x_continuous(breaks = seq(0, 1, by = 0.1)) -->
<!-- cmps.252.plot <- metric_plot_helper(CMPS_hamby252_results_npeak$cmps.table[[8]], "cmps.diff_scaled",  -->
<!--                                     scaled = TRUE, subtitle = "CMPS, 10-8-6-4-2-1, 50") + -->
<!--   xlab(labels[3]) + -->
<!--   scale_x_continuous(breaks = seq(0, 1, by = 0.1)) -->

<!-- rf.44.plot <- metric_plot_helper(rf.results[[2]], "rf.diff", scaled = TRUE, -->
<!--                                   subtitle = "random forest scores with metric diff") + -->
<!--   xlab(expression(RF[diff])) + -->
<!--   scale_x_continuous(breaks = seq(0, 1, by = 0.1)) -->
<!-- cmps.44.plot <- metric_plot_helper(CMPS_hamby44_results_npeak$cmps.table[[6]], "cmps.diff_scaled",  -->
<!--                                     scaled = TRUE, subtitle = "CMPS, 10-7-4-2, 61") + -->
<!--   xlab(labels[3]) + -->
<!--   scale_x_continuous(breaks = seq(0, 1, by = 0.1)) -->

<!-- p.thisisus252 <- ggarrange(plotlist = list(cmps.252.plot, rf.252.plot), nrow = 1, ncol = 2, -->
<!--                common.legend = TRUE, legend = "bottom") -->
<!-- p.thisisus252 <-  -->
<!--   annotate_figure(p.thisisus252,  -->
<!--                   top = text_grob("Hamby 252 - Random Forest and CMPS")) -->

<!-- p.thisisus44 <- ggarrange(plotlist = list(cmps.44.plot, rf.44.plot), nrow = 1, ncol = 2, -->
<!--                common.legend = TRUE, legend = "bottom") -->
<!-- p.thisisus44 <-  -->
<!--   annotate_figure(p.thisisus44,  -->
<!--                   top = text_grob("Hamby 44 - Random Forest and CMPS")) -->

<!-- ``` -->

>>>>

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
library(tidyverse)
library(patchwork)
library(ggpubr)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "\\textwidth",
  fig.align = "center",
#  cache=TRUE,
  dpi = 144,
  warning = FALSE,
  message = FALSE
)

@@

```{r func, echo=FALSE}
bootstrap_k <- function(scores, k, K, value) {
  res <- replicate(K, {
    mean(sample(scores, size = k, replace = TRUE), na.rm = TRUE)
  })
  sum(res >= value) / K
}

compute_average_scores <- function(land1, land2, score, addNA = FALSE) {
  if (!is.numeric(land1)) land1 <- readr::parse_number(as.character(land1))
  if (!is.numeric(land2)) land2 <- readr::parse_number(as.character(land2))
  # assert_that(is.numeric(land1), is.numeric(land2), is.numeric(score))

  maxland <- max(land1, land2)
  fullframe <- data.frame(expand.grid(land1 = 1:maxland, land2 = 1:maxland))
  bcompare <- data.frame(land1, land2, score)

  fullframe <- fullframe %>% left_join(bcompare, by = c("land1", "land2"))

  fullframe <- fullframe %>% mutate(
    land1 = factor(land1, levels = 1:maxland),
    land2 = factor(land2, levels = 1:maxland)
  )
  # get averages, just in case
  matrix <- xtabs(score ~ land1 + land2,
    data = fullframe, addNA = addNA
  ) / xtabs(~land1 + land2, data = fullframe, addNA = addNA)

  matrix <- cbind(matrix, matrix)

  scores <- 1:maxland %>% sapply(FUN = function(i) {
    if (i == 1) {
      mean(diag(matrix), na.rm = TRUE)
    } else {
      i <- i - 1
      mean(diag(matrix[, -(1:i)]), na.rm = TRUE)
    }
  })
  scores
}

bullet_to_land_predict <- function(land1, land2, scores, difference, alpha = 0.05, addNA = FALSE) {
  if (!is.numeric(land1)) land1 <- readr::parse_number(as.character(land1))
  if (!is.numeric(land2)) land2 <- readr::parse_number(as.character(land2))
  avgs <- compute_average_scores(land2, land1, scores, addNA)

  p <- max(c(land1, land2))
  boot <- bootstrap_k(scores, p, 1000, max(avgs)) < alpha
  if (!is.numeric(boot)) {
    boot <- bootstrap_k(scores, p, 1000, max(avgs)) < alpha
    #     print("boot is not numeric")
    #      browser()
  }

  getdiff <- diff(sort(-avgs))[1]
  if (!is.numeric(getdiff)) print("getdiff is not numeric") # browser()
  if (getdiff > difference & boot) {
    # pick the maximum to determine the phase
    idx <- which.max(avgs)
    dd <- data.frame(
      land1,
      land2
    ) %>%
      mutate_if(function(.) !is.numeric(.), parse_number)
    dd$diff <- (dd$land1 - dd$land2) %% p + 1

    return(dd$diff == idx)
  } else {
    return(rep(FALSE, length = length(land1)))
  }
}
```

@@

In this paper we present an open-source implementation of the algorithm of the Congruent Matching Profile Segments (CMPS) method.
@cmps developed the CMPS method for "objective comparison of striated tool marks" and demonstrated its use in some examples of comparing bullet signature correlations.
Although @cmps conceptually describe the CMPS algorithm in their paper, the authors did not release an implementation of their method.
Thus, our effort here is to introduce the CMPS method to the `R` [@R] community and provide an open-source, publicly available implementation of the algorithm to use, review, and improve.
Our implementation is made available as part of the `R` package \pkg{cmpsR} on CRAN.

According to the Uniform Crime Reporting Program of FBI (Federal Bureau of Investigation), "more than 76 percent (76.7) of the homicides for which the FBI received weapons data in 2020 involved the use of firearms" [@fbiucr]. At the same time, the number of murder victims jumped between 2019 and 2020 by more than 23 percent (23.4) to 17754 [@fbiucr20]. This increase is unprecedented  and highlights the important role that  firearm examination plays in all these cases. 
<!-- homicides in 2019: 14391; in 2020: 17754 -->
An important task of firearm examination is to answer the question of whether two pieces of evidence come from the same source or whether a piece of evidence matches a sample obtained from a specific firearm.
Here, in particular, we are interested to determine whether two bullets were fired from the same gun barrel.
Assessing the similarity between two bullets is based on a comparison of striation marks acquired during the firing process as bullets are propelled through the barrel.
Current state of art sees firearms examiners make an assessment of similarity based on a visual comparison, generally, using a comparison microscope [@afte].
This practice has been criticized for its lack of objectivity and the associated problem of determining valid error rates [@pcast].

@@

The datasets we worked with come from the James Hamby Consecutively Rifled Ruger Barrel Study [@brundage; @hamby; @Hamby:2019], in particular, Hamby set 252 and Hamby set 44.
For each Hamby set, a total of 35 bullets is fired from (the same) ten consecutively manufactured Ruger P-85 pistol barrels.
Two bullets are fired from each barrel, making up a set of 20 reference bullets.
An additional 15 bullets are fired from these ten barrels in a fashion unknown to the study participant.
The aim of the Hamby Study was to have firearms examiners identify which barrel each of the 15 questioned bullets was fired from.
The Ruger P-85 barrels are traditionally rifled barrels with six grooves and lands as shown in \autoref{fig:bullet}.
During the firing process, grooves and lands are engraved on a bullet.
Firearms examiners use striation marks on land engraved areas (LEAs) for their visual comparison.
For algorithmic purposes, 3D topographical images of land engraved areas were obtained and stored in x3p format (XML 3-D Surface Profile).
The x3p format provides a standard way of exchanging 2D and 3D profile data.
It conforms to [the ISO5436-2 standard](http://sourceforge.net/p/open-gps/mwiki/X3p/) adopted by the OpenFMC (Open Forensic Metrology Consortium), a group of firearm forensics researchers who contributes to the establishment of best practices of using metrology in forensic science.
Hamby set 252 was scanned using a NanoFocus lens at 20x magnification with the scan resolution being 1.5625 \textmu m $\times$ 1.5625 \textmu m per pixel.
Hamby set 44 was scanned at the Roy J Carver High-Resolution microscopy lab at Iowa State.
These scans were acquired with a Sensofar Confocal Light Microscope at 20x magnification for a nominal resolution of 0.645 \textmu m $\times$ 0.645 \textmu m per pixel.
Both Hamby set 252 and Hamby set 44 are publicly available from the NIST Ballistics Database Project [@nistdb].

@@

The thin white horizontal line in \autoref{fig:process}(a) indicates which cross-section was identified by the algorithm to represent the LEA; \autoref{fig:process}(b) shows the corresponding cross-sectional view.
Groove engraved areas are removed from the analysis as indicated by the vertical blue lines in \autoref{fig:process}(c) and \autoref{fig:process}(d)).
A non-parametric LOESS smooth [@loess] is fitted to capture the bullet curvature (\autoref{fig:process}(e)) and, finally, the **bullet signature** (\autoref{fig:process}(f)) is obtained as residuals of the cross-section and the fitted smooth.
Note, that in @cmps bullet signatures are referred to as bullet profiles.
However, to avoid confusion, we distinguish the notion of bullet signatures from bullet profiles.
Bullet profiles are shown in panels (b), (c), and (d) of \autoref{fig:process}, while \autoref{fig:process}(f) shows the corresponding bullet signature.
Identifying the groove engraved areas correctly is fundamental for a correct down-stream analysis of the signatures.
To allow for a human-in-the middle inspection and intervention we have provided an interactive web-application, implemented as `R` Shiny App [@shiny], \pkg{bulletinspectR} to identify and correct those errors.
An example of the extraction process with corresponding code and parameter settings can be found in the "Supplementary materials". Note that the process of extracting signatures might be different from the one used in @cmps because no code or parameter settings are made available publicly.

@@

Most algorithms for comparing striation marks are based on the digitized signatures and produce a similarity score [@song2005, @ChumbleyL_Scott2010VoTM, @aoas, @pmid30444940].
The congruent matching profile segments (CMPS) algorithm, developed by @cmps for "objective comparison of striated tool marks", is one such algorithm.
The algorithm's main idea is to take a set of consecutive and non-overlapping basis segments from the comparison and for each segment find the "best" registration position on the reference (the other bullet signature) with respect to their cross-correlation values.
From a comparison of these registration positions, a **congruent registration position** is identified, and the number of basis segments taking the congruent registration position is the CMPS score.
Note that researchers in @cmps refer to the origin of basis segments as the reference, but in this paper we refer to it as the comparison signature.
High CMPS scores are achieved between more similar signatures and are therefore indicative of a same-source pair.
Low scores between pairs of signatures are attributed to different source pairs.
However, a specific threshold of the CMPS score to distinguish between same-source and different-source comparisons is not provided in @cmps, instead the threshold depends on the underlying structure of the data and the choice of parameters.   In a legal setting this variability is problematic because it allows for situations in which experts could choose parameters based on whether they are witnesses for the defense or the prosecution. 
Further research is needed to understand how to determine optimal threshold settings.
The CMPS algorithm can assist firearm examiners with drawing a conclusion about the source of a comparison pair.
Thus, in this paper we present an open-source implementation of the CMPS algorithm in the R package \pkg{cmpsR} available from both CRAN and Github. This publicly available implementation calculates the CMPS score of a comparison using the following code:

@@

@@

@@

For details about the implementation, examples and results, please refer to the "Supplementary materials".

@@

**Data not included in "supplementary-files.zip"**

@@

**Data included in "supplementary-files.zip"**

-   `data-csv`: this folder contains `.csv` files we generated using R scripts `hamby*_result_generator.R` 
-   `CMPSpaper_results.rds`: this is the `.rds` file we generated using R script `rds_generator.R`

These data can be used as reference or example results of the reproducible codes

-   `csafe_rf2.rds`: this `.rds` file contains the random forest model used in this paper
